---
title: "Angles"
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
format: 
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{angles}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r presetup, include = FALSE}
library(ggpathdiagramr)
library(ggplot2)
library(dplyr)
library(ggtext)
library(ggarrow)
knitr::opts_chunk$set(dev = "svglite", fig.height = 8, fig.width = 8)
my_font <- "Roboto Condensed"
gdtools::register_gfont(family = my_font)
gdtools::addGFontHtmlDependency(family = my_font)
update_geom_defaults("richtext",
                     list(family = my_font, 
                          size = .8 * 16 / ggplot2::.pt))

update_geom_defaults("arrow_segment", list(length_head = 1))
```

```{r setup}
library(ggpathdiagramr)
library(ggplot2)
library(dplyr)
library(ggtext)
library(ggarrow)
my_font <- "Roboto Condensed"

```


Angles have different kinds of units associated with them: degrees, radians, and turns. The `angle` class can take turns (i.e., 1 turn = one full rotation a circle), degrees (1 turn = 360 degrees), or radians (1 turn = $2\pi$ = $\tau$).

I like &pi; just fine, but I agree with Michael Hartl's [Tau Manifesto](https://tauday.com/tau-manifesto) that we would have been better off if we had recognized that the number of radians to complete a full turn of a circle (&tau; = 2&pi; &asymp; 6.283185) is more fundamental than the number of radians to complete a half turn (&pi;). 

Turns | Radians | Degrees |
:----:|:-------:|:-------:|
$\frac{1}{12}$  | $\frac{\tau}{12}=\frac{\pi}{6}$   | $30^\circ$  |
$\frac{1}{8}$  | $\frac{\tau}{8}=\frac{\pi}{4}$     | $45^\circ$  |
$\frac{1}{6}$  | $\frac{\tau}{6}=\frac{\pi}{3}$     | $60^\circ$  |
$\frac{1}{4}$  | $\frac{\tau}{4}=\frac{\pi}{2}$     | $90^\circ$  |
$\frac{1}{3}$  | $\frac{\tau}{3}=\frac{2\pi}{3}$    | $120^\circ$ |
$\frac{1}{2}$  | $\frac{\tau}{2}=\pi$               | $180^\circ$ |
$1$            | $\tau=2\pi$                        | $360^\circ$ |
