---
title: "Lines"
format: 
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{lines}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: presetup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  dev = "ragg_png",
  comment = "#>",
  fig.width = 7,
  fig.height = 7
)
library(ggdiagram)
library(ggplot2)
library(dplyr)
library(ggtext)
library(ggarrow)
my_font <- "Roboto Condensed"
```

# Setup

## Packages

```{r}
#| label: setup
library(ggdiagram)
library(ggplot2)
library(dplyr)
library(ggtext)
library(ggarrow)
```

## Base plot

To avoid repetitive code, we make a base plot:

```{r}
#| label: baseplot
my_font <- "Roboto Condensed"
my_font_size <- 20
my_point_size <- 2.5
# my_colors <- viridis::viridis(2, begin = .25, end = .5)
my_colors <- c("#3B528B", "#21908C")
update_geom_defaults(geom = "richtext", 
                     new = list(family = my_font, 
                                size = .8 * my_font_size / ggplot2::.pt))

update_geom_defaults("point", list(size = my_point_size))
update_geom_defaults("arrow_segment", list(length_head = 5, linewidth = .5))

theme_set(
  theme_minimal(
    base_size = my_font_size, 
    base_family = my_font) +
    theme(axis.title.y = element_text(angle = 0, vjust = 0.5)))




bp <- ggplot() +
  coord_equal() +
  scale_x_continuous(labels = WJSmisc::signs_centered, limits = c(-4, 4)) +
  scale_y_continuous(labels = signs::signs, limits = c(-4, 4)) +
  theme_minimal(base_size = 16) +
  theme(
    axis.title.x =  element_text(face = "italic"),
    axis.title.y = element_text(
      face = "italic",
      angle = 0,
      hjust = .5,
      vjust = .5
    )
  )
```

# Making Lines

Lines can be constructed from a slope and an intercept:

```{r}
#| label: construct_line
l <- ob_line(slope = 2, intercept = 1, color = my_colors[1])
l
```

```{r}
#| label: fig-slopeintercept
#| fig-cap: !expr 'paste0("A line with slope of ", l@slope, " and intercept of ", l@intercept)'
#| code-fold: true

bp +
  l +
  l@point_at_y(0)@label(l@equation, angle = l@angle)
```


Because the default slope is 0, a horizontal ob_line can be set with just the intercept:

```{r}
#| label: hline
h <- ob_line(intercept = 2, color = my_colors[1])
h
```

```{r}
#| label: fig-horitonal
#| fig-cap: !expr 'paste0("A horizontal line intercept of ", h@intercept)'
#| code-fold: true
bp + 
  h + 
  h@point_at_x(0)@label(h@equation)
```


A vertical line can be set with the x-intercept:

```{r}
#| label: vline
v <- ob_line(xintercept = -2, color = my_colors[1])
v
```

```{r}
#| label: fig-v
#| fig-cap: !expr 'paste0("Vertical line at x = ", signs::signs(v@xintercept))'
#| code-fold: true
bp + 
  v + 
  v@point_at_y(0)@label(equation(v), angle = v@angle * -1)

```



Any line---horizontal, vertical, or sloped---can be constructed from the coefficients of the general form of a line:

$$
ax+by+c=0
$$

```{r}
#| label: abcline
l_123 <- ob_line(a = 1, b = 2, c = 3, color = my_colors[1])
```

```{r}
#| label: fig-ablcline_plot
#| fig-cap: !expr 'paste0("Line with slope = ", signs::signs(l_123@slope), " and intercept = ", signs::signs(l_123@intercept))'
#| code-fold: true
bp +
  l_123 +
  l_123@point_at_x(
    x = 0)@label(
      equation(l_123, type = "general"), 
      angle = l_123@angle)
```

With respect to the general form, the slope is equal to $-\frac{a}{b}$, the y-intercept is equal to $-\frac{c}{b}$, and the x-intercept is equal to $-\frac{c}{a}$

# Methods

## Projections and Distances

A point can be "projected" onto a line. Imagine shining a light on the point in a direction perpendicular to the line. The point's shadow on the line would be the shortest distance between the line and the point.


```{r}
#| label: projection
p <- ob_point(3,-2, color = my_colors[2])
l <- ob_line(slope = 2, intercept = 2, color = my_colors[1])
# Point p projected onto line l
p_projected <- projection(p, l)

# Alternately:
l@projection(p)
```


The shortest distance from a point to a line can be calculated.

```{r}
#| label: short
# distance from point p to line l
distance(p, l)

# Equivalently:
ob_segment(p, l@projection(p))@distance
```

```{r}
#| label: fig-point_line
#| fig-cap: Shortest distance between a line and point
#| code-fold: true

bp +
  l +
  l@point_at_x(.5)@label(
    label = l@equation, 
    angle = l@angle) +
  {s_projected <- ob_segment(
      p1 = l@projection(p),
      p2 = p,
      linetype = "dashed",
      label = paste0("Distance = ", 
                     distance(l@projection(p), p) |>
                       round(digits = 2) |>
                       as.character()))} + 
  s_projected@midpoint(c(0, 1))@label(
    polar_just = degree(s_projected@line@angle) + c(180, 0),
    plot_point = TRUE)  


```



