---
title: "Paths"
format: 
  html:
    toc: true
    html-math-method: katex
vignette: >
  %\VignetteIndexEntry{paths}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: presetup
#| include: false
library(ggdiagram)
library(ggplot2)
library(dplyr)
library(ggtext)
library(ggarrow)
knitr::opts_chunk$set(
  collapse = TRUE,
  dev = "ragg_png",
  comment = "#>",
  fig.width = 7,
  fig.height = 7
)

```

# Setup

## Packages

```{r}
#| label: setup
library(ggdiagram)
library(ggplot2)
library(dplyr)
library(ggtext)
library(ggarrow)
```


## Base Plot

To avoid repetitive code, we make a base plot:

```{r}
#| label: baseplot
#| file: baseplot.R
```

# Paths

The `path` function creates an object that connects points along a path.

```{r}
#| label: fig-path
#| fig-cap: Plotting a path.
p <- ob_point(c(-4,-3,4), c(0,-2, 0), color = "dodgerblue")
bp + 
  ob_path(p) +
  p
```


# Path Labels

The label of a path is created with `geomtextpath::geom_labelpath`, and thus will curve if the path is curved.


```{r}
#| label: fig-pathlabel
#| fig-cap: A path with a curved label
p_curve <- tibble(x = seq(-4, 4, .1), 
                  y = (pnorm(x) * 8 - 4)) |>
  ob_point()



bp +
  ob_path(
    p = p_curve,
    label = ob_label(
      "A path label will curve along the path.",
      vjust = -.1,
      size = 20,
      color = "dodgerblue4"
    ),
    arrowhead_length = 8,
    arrow_head = my_arrowhead
  )


```

You can control the position of the path label with either the label's `position` or `hjust` properties. 

```{r}
#| label: fig-labelposition
#| fig-cap: A path with a curved label at position .1
bp +
  ob_path(
    p = p_curve,
    label = ob_label(
      "A path label will curve along the path.",
      vjust = -.1,
      size = 20,
      color = "dodgerblue4",
      position = .1
    ),
    arrowhead_length = 8,
    arrow_head = my_arrowhead
  )
```


# Multiple paths

To create multiple paths at once, specify a list or vector of point objects.

```{r}
#| label: fig-multiplepath
#| fig-cap: Multiple paths
bp +
  ob_path(c(p_curve, 
         p_curve * .5 + ob_point(2,-2),
         p_curve * .5 + ob_point(-2,2)), 
       color = c("orchid4",
                 "dodgerblue3",
                 "firebrick4"),
       label = c(".52", ".42", ".61"))
```

