---
title: "Polygons"
format: 
  html:
    toc: true
    html-math-method: katex
vignette: >
  %\VignetteIndexEntry{polygons}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: presetup
#| include: false
library(ggdiagram)
library(ggplot2)
library(dplyr)
library(ggtext)
library(ggarrow)
knitr::opts_chunk$set(
  collapse = TRUE,
  dev = "ragg_png",
  comment = "#>",
  fig.width = 7,
  fig.height = 7
)

```

# Setup

## Packages

```{r}
#| label: setup
library(ggdiagram)
library(ggplot2)
library(dplyr)
library(ggtext)
library(ggarrow)
library(arrowheadr)

```

## Base Plot

To avoid repetitive code, we make a base plot:

```{r}
#| label: baseplot


my_font <- "Roboto Condensed"
my_font_size <- 20
my_point_size <- 2
my_arrowhead <- arrowheadr::arrow_head_deltoid(2.1)

# my_colors <- viridis::viridis(2, begin = .25, end = .5)
my_colors <- c("#3B528B", "#21908C")

theme_set(
  theme_minimal(
    base_size = my_font_size,
    base_family = my_font) +
    theme(axis.title.y = element_text(angle = 0, vjust = 0.5)))

bp <- ggdiagram(
  font_family = my_font,
  font_size = my_font_size,
  point_size = my_point_size,
  linewidth = .5,
  theme_function = theme_minimal,
  axis.title.x =  element_text(face = "italic"),
  axis.title.y = element_text(
    face = "italic",
    angle = 0,
    hjust = .5,
    vjust = .5)) +
  scale_x_continuous(labels = WJSmisc::signs_centered,
                     limits = c(-4, 4)) +
  scale_y_continuous(labels = signs::signs,
                     limits = c(-4, 4))

```

# Polygons

The `ob_polygon` function creates an object that connects points to make a polygon.

```{r}
#| label: fig-path
#| fig-cap: Plotting a path.
p <- ob_point(x = c(-4,-3, 4,  3, -1), 
           y = c( 0, 4, 3, -3, -4))
bp + 
  ob_polygon(p, fill = "dodgerblue3")
```


# Polygon Labels

The label of a `ob_polygon` object is placed, by default, at the centroid of the polygon. The centroid is the point whose x coordinate is the average of all the x coordinates of the polygon's points and whose y coordinate is likewise the average of all the point's y coordinates.


```{r}
#| label: fig-polygonlabel
#| fig-cap: A path with a curved label
bp +
  ob_polygon(
    p = p,
    label = ob_label(
      "My<br>Polygon",
      size = 20,
      color = "white"
    ),
    fill = "dodgerblue4"
  )

```

# Connecting polygons

Connections between polygons are arrows that emanate from and point towards the polygons' centroids.

```{r}
#| label: fig-polygonconnect
#| fig-cap: Arrow between two polygons
bp + 
  {p1 <- ob_polygon(.5 * p - ob_point(1.9,1.9), 
              fill = "dodgerblue4")} + 
  {p2 <- ob_polygon(.5 * p + ob_point(1.9, 1.9), 
              fill = "orchid4")} + 
  connect(p1, p2, 
          resect = 1, 
          label = ".45")
```

# Rounding polygons

The `@radius` property controls the radius of the rounded vertices. It must be of length 1. It can be given in as a `ggplot2::unit` or as a numeric value. If numeric, it is understood as a proportion of the plot area width.

```{r}
#| label: fig-roundedpolygon
#| fig-cap: A polygon with rounded vertices
bp + 
  ob_polygon(
    p, 
    radius = unit(5, "mm"),
    fill = "dodgerblue4")
```

