---
title: "Rectangles"
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
format: 
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{rectangles}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r presetup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  dev = "ragg_png",
  comment = "#>",
  fig.width = 7,
  fig.height = 7
)

library(ggdiagram)
library(ggplot2)
library(ggtext)
library(ggarrow)
library(arrowheadr)

my_font <- "Roboto Condensed"
# gdtools::register_gfont(family = my_font)
# gdtools::addGFontHtmlDependency(family = my_font)
```

```{r setup}
library(ggdiagram)
library(ggplot2)
library(ggtext)
library(ggarrow)
library(arrowheadr)
my_font <- "Roboto Condensed"
my_text_size <- 20
my_point_size <- 3
my_colors <- list(
  primary = class_color("royalblue4"),
  secondary = class_color("firebrick4"),
  tertiary = class_color("orchid4"))
update_geom_defaults(
  geom = "richtext", 
  new = list(family = my_font, 
             size = .8 * my_text_size / ggplot2::.pt))

update_geom_defaults(
  geom = "point", 
  list(size = my_point_size))


theme_set(
  theme_minimal(
    base_size = my_text_size, 
    base_family = my_font) +
    theme(
      axis.title.y = element_text(angle = 0, vjust = 0.5, face = "italic"),
      axis.title.x = element_text(face = "italic")))
```

# Specifying a Rectangle

A rectangle has 4 corners (`northeast`, `northwest`, `southwest`, and `southeast`). It has a center. It has width and height. For the purpose of demonstration, we can specify all these features, though in practice not all of them are necessary.

```{r corners}
# northeast corner
ne <- point(4,2) 
# northwest corner
nw <- point(0,2)
# southwest corner
sw <- point(0,0)
# southeast corner
se <- point(4,0)
# centner point
cent <- point(2,1)
# width
w <- 4
# height
h <- 2
```


If you give the `rectangle` function enough information to deduce where its four corners will be, all other features will be calculated. All of the following will give the same rectangle:

## Give width, height, and any point

An easy way to specify a rectangle is to specify its width and height and any of its points. All the following rectangles are equivalent.

### Center, width, and height


```{r centerhw}
r1 <- rectangle(
  width = w,
  height = h,
  center = cent,
  color = my_colors$primary,
  fill = my_colors$primary@transparentize(.15),
  linewidth = 1
)
r1
```



```{r fig-rectangle}
#| label: fig-angles
#| code-fold: true
#| fig-cap: "Center, Width, and Height"
#| fig-width: 8
#| fig-height: 4.75
double_arrowstyle <- style(
  arrow_head = arrow_head_deltoid(),
  arrow_fins = arrow_head_deltoid(),
  color = my_colors$secondar
)

s_east <- r1@side@east@nudge(x = .1)
s_east@style <- double_arrowstyle

s_north <- r1@side@north@nudge(y = .1)
s_north@style <- double_arrowstyle

rc_plot <- ggplot() +
  coord_equal(ylim = c(0, 2.2)) +
  scale_y_continuous(breaks = -10:10) +
  r1

rc_center <- c(
  as.geom(r1@center),
  as.geom(r1@center@label(
    fill = my_colors$primary@lighten(.15), 
    vjust = -.15)))

rc_width <- c(
  as.geom(s_north), 
  as.geom(s_north@label(
    label = "Width = " + r1@width,
    color = my_colors$secondary,
    vjust = 0,
    label.margin = ggplot2::margin(2, 2, 2, 2, "pt")
  )))

rc_height <- c(
  as.geom(s_east),
  as.geom(s_east@label(
    label = "Height = " + r1@height,
    vjust = 0,
    color = my_colors$secondary,
    angle = -90)
  )
)

rc_nw <- r1@northwest@label(
  plot_point = T,
  vjust = 1.1,
  hjust = 0,
  fill = my_colors$primary@lighten(.15)
)

rc_ne <- r1@northeast@label(
  plot_point = T,
  vjust = 1.1,
  hjust = 1,
  fill = my_colors$primary@lighten(.15)
)

rc_sw <- r1@southwest@label(
  plot_point = T,
  vjust = -.1,
  hjust = 0,
  fill = my_colors$primary@lighten(.15)
)

rc_se <- r1@southeast@label(
  plot_point = T,
  vjust = -.1,
  hjust = 1,
  fill = my_colors$primary@lighten(.15)
)

rc_plot + rc_center + rc_width + rc_height
```



### Northeast corner, width, and height

```{r hw_ne}
r1 == rectangle(width = w,
                height = h,
                northeast = ne)
```

```{r fig-hw_ne}
#| fig-cap: Specify a rectangle with height, width, and the northeast corner
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_width + rc_height + rc_ne
```

### Northwest corner, width, and height

```{r hw_nw}
r1 == rectangle(width = w,
                height = h,
                northwest = nw)
```


```{r fig-hw_nw}
#| fig-cap: Specify a rectangle with height, width, and the northwest corner
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_width + rc_height + rc_nw
```

### Southwest corner, width, and height


```{r hw_sw}
r1 == rectangle(width = w,
                height = h,
                southwest = sw)
```


```{r fig-hw_sw}
#| fig-cap: Specify a rectangle with height, width, and the southwest corner
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_width + rc_height + rc_sw
```

### Southeast corner, width, and height

```{r hw_se}
r1 == rectangle(width = w,
                height = h,
                southeast = se)
```


```{r fig-hw_se}
#| fig-cap: Specify a rectangle with height, width, and the southeast corner
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_width + rc_height + rc_se
```




## Give the center and any of the 4 corners

A rectangle can be specified with the center and any other corner. The following rectangles are equivalent.

### Northeast corner and center

```{r center_ne}
r1 == rectangle(center = cent, 
                northeast = ne)
```

```{r fig-center_ne}
#| fig-cap: Specify a rectangle with the center and the northeast corner
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_center + rc_ne
```

### Northwest corner and center

```{r center_nw}
r1 == rectangle(center = cent, 
                northwest = nw)
```

```{r fig-center_nw}
#| fig-cap: Specify a rectangle with the center and the northwest corner
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_center + rc_nw
```

### Southeast corner and center

```{r center_se}
r1 == rectangle(center = cent, 
                southeast = se)
```

```{r fig-center_se}
#| fig-cap: Specify a rectangle with the center and the southeast corner
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_center + rc_se
```

### Southeast corner and center

```{r center_sw}
r1 == rectangle(center = cent, 
                southwest = sw)
```

```{r fig-center_sw}
#| fig-cap: Specify a rectangle with the center and the southwest corner
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_center + rc_sw
```


## Give opposite corners

A rectangle can be specified with points from opposite corners. These rectangles are equivalent.

### Northeast and southwest corners

```{r ne_sw}
r1 == rectangle(
  northeast = ne, 
  southwest = sw)
```

```{r fig-ne_sw}
#| fig-cap: Specify a rectangle with the northeast and southwest corners
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_sw + rc_ne
```

### Northwest and southeast corners

```{r nw_se}
r1 == rectangle(
  northeast = ne, 
  southwest = sw)
```

```{r fig-nw_se}
#| fig-cap: Specify a rectangle with the northwest and southeast corners
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_se + rc_nw
```

## Give width and two points on either side

A rectangle can be specified with the width and 2 points from the left or right side. These rectangles are equivalent.

```{r w_nw_sw}
r1 == rectangle(width = w,
                northwest = nw,
                southwest = sw)
```

```{r fig-w_nw_sw}
#| fig-cap: Specify a rectangle with the width and the left corners
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_width + rc_nw + rc_sw
```

```{r w_ne_se}
r1 == rectangle(width = w,
                northwest = ne,
                southwest = se)
```

```{r fig-w_ne_se}
#| fig-cap: Specify a rectangle with the width and the right corners
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_width + rc_ne + rc_se
```

## Give height and two points on top or bottom

A rectangle can be specified with the height and 2 points from the top or bottom side. These rectangles are equivalent.

## Height and top corners

```{r h_ne_nw}
r1 == rectangle(height = h,
                northwest = nw,
                northeast = ne)
```

```{r fig-h_ne_nw}
#| fig-cap: Specify a rectangle with the height and the top corners
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_height + rc_ne + rc_nw
```

## Height and bottom corners

```{r n_sw_se}
r1 == rectangle(height = h,
                southwest = sw,
                southeast = se)
```

```{r fig-n_sw_se}
#| fig-cap: Specify a rectangle with the height and the top corners
#| code-fold: true
#| fig-height: 4.75
rc_plot + rc_sw + rc_se + rc_height
```


# Rectangle points

The corners and side midpoints can be extracted. Here is the north point (i.e., the midpoint of the north side):

```{r north}
r1@north
```

```{r fig-points}
#| fig-cap: The named points of a rectangle
#| code-fold: true
#| fig-height: 4.75

rc_plot +
  purrr::map(
    c(
      "east",
      "north",
      "west",
      "south",
      "northeast",
      "northwest",
      "southeast",
      "southwest",
      "center"
    ),
    \(x) {
      v <- ifelse(grepl(x = x, "north"),
        1.1,
        ifelse(grepl(x = x, "south|center"), 
               -.1, 
               .5))
      h <- ifelse(grepl(x = x, "east"),
                  1.1,
                  ifelse(grepl(x = x, "west"), -.1, .5))
      c(
        as.geom(
          prop(r1, x)@label(
            label = x,
            hjust = h,
            vjust = v,
            fill = my_colors$primary@lighten(.15)
          )
        ),
        as.geom(
          prop(r1, x)@label(hjust = 1 - h, vjust = 1 - v),
          fill = ifelse(x == "center", 
                        my_colors$primary@lighten(.15), 
                        "white")
        ),
        as.geom(prop(r1, x))
      )
    }
  ) +
  coord_equal(xlim = c(-.25, 4.25), ylim = c(-.25, 2.25))
  

```

Points at any angle

```{r theta60}
theta <- degree(60)
r1@point_at_theta(theta)
```


```{r fig-theta60}
#| fig-cap: "The point on a rectangle at angle &theta; = 60&deg;"
#| code-fold: true
#| fig-height: 4.75
r1_theta <- r1@point_at_theta(theta)

rc_plot +
  segment(r1@center, r1_theta) +
  r1_theta@label(
    polar_just = polar(theta, 1.5), 
    plot_point = T) +
  arc(center = r1@center,
      radius = .5,
      end = theta) +
  arc(center = r1@center,
      radius = .5,
      end = theta)@label(fill = my_colors$primary@lighten(.15)) 
```



# Rectangle sides

Each side of the rectangle can be extracted. For example, here is the north side segment:

```{r westside}
r1@side@north
```

```{r fig-westside}
#| fig-cap: The north side of a rectangle
#| code-fold: true
#| fig-height: 4.75
rc_plot + 
  r1@side@north %>% 
    set_props(color = my_colors$secondary@color, linewidth = 2) +
  r1@north@label(label = "North Side",
                 vjust = -.1, 
                 size = 20,
                 color = my_colors$secondary)
```



