---
title: "Segments"
format: 
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{segments}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r presetup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dev = "ragg_png",
  comment = "#>",
  fig.width = 7,
  fig.height = 7
)
library(ggdiagram)
library(ggplot2)
library(ggtext)
```

```{r setup}
library(ggdiagram)
library(ggplot2)
library(ggtext)
```

For ease of plotting, we will define a base plot:

```{r}
my_font <- "Roboto Condensed"
update_geom_defaults("richtext", 
                     list(family = my_font, 
                          size = 0.8 * 20 / ggplot2::.pt))
update_geom_defaults("line", list(linewidth = .75))
update_geom_defaults("segment", list(linewidth = .75))
update_geom_defaults("point", list(size = 2))

bp <- ggplot() +
  theme_minimal(base_size = 20, base_family = my_font) +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) +
  coord_equal() +
  xlim(-4,4) +
  ylim(-4,4)


```


A segment is a portion of a line between two points.


```{r}
p1 <- point(-3,-1)
p2 <- point(3,3)
s1 <- segment(p1, p2)
```

```{r}
bp + s1 + p1 + p2
```

Styles can be specified when the segment is created.

```{r}
s2 <- segment(p1,p2, color = "red")
bp + s2
```

Styles can be modified via the `@style` property:

```{r}
s2@style@linewidth <- 3
bp + s2
```

The `as.geom` passes style arguments to the `geom_seqment` function without modifying the segment's style property:

```{r}
# Make s1 appear red temporarily
bp +
  as.geom(s1, color = "red")
# Verify that s1's style is not red
bp + s1
```

A "pipe-friendly" way to modify any ggdiagram object is to use S7's `set_prop` function, which has been re-exported to ggdiagram for the sake of convenience. 

```{r}

bp +
  s1 %>% 
    set_props(color = "red") 

```



The line that passes between both points:

```{r}
s1@line
```


```{r}
s1_ends <- midpoint(s1, position = c(0,1))
bp +
  s1@line +
  s1@midpoint(position = c(0,1))@label(
    polar_just = polar(s1@line@angle + degree(90), 1.5), 
    plot_point = TRUE) +
  s1@label(equation(s1@line))
```

The distance between a segment's points:

```{r}
distance(s1)
```

```{r}
bp + 
  s1 + 
  s1@label(distance(s1)) 
```


