---
title: "Segments"
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
    dev: ragg_png
format: 
  html:
    toc: true
    html-math-method: katex
    fig-width:  7
    fig-height: 7
vignette: >
  %\VignetteIndexEntry{segments}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: presetup
#| include: false

library(ggdiagram)
library(ggplot2)
library(ggtext)
```

# Setup

## Packages

```{r}
#| label: setup
library(ggdiagram)
library(ggplot2)
library(dplyr)
library(ggtext)
library(ggarrow)
library(arrowheadr)

```

## Base Plot

To avoid repetitive code, we make a base plot:

```{r}
#| label: baseplot
my_font <- "Roboto Condensed"
my_font_size <- 20
my_point_size <- 2

# my_colors <- viridis::viridis(2, begin = .25, end = .5)
my_colors <- c("#3B528B", "#21908C")

theme_set(
  theme_minimal(base_size = my_font_size, base_family = my_font) +
    theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
)

bp <- ggdiagram(
  font_family = my_font,
  font_size = my_font_size,
  point_size = my_point_size,
  linewidth = .5,
  theme_function = theme_minimal,
  axis.title.x =  element_text(face = "italic"),
  axis.title.y = element_text(
    face = "italic",
    angle = 0,
    hjust = .5,
    vjust = .5
  )
) +
  scale_x_continuous(labels = signs_centered, limits = c(-4, 4)) +
  scale_y_continuous(labels = signs::signs, limits = c(-4, 4))
```

# Specifying a segment

A segment is a portion of a line between two points.


```{r}
#| label: s1
p1 <- ob_point(-3, -1)
p2 <- ob_point(3, 3)
s1 <- ob_segment(p1, p2)
```

```{r}
#| label: fig-s1
#| fig-cap: Plotting a segment and its endpoints
bp + s1 + p1 + p2
```



# Features of a segment

## Distance between points

```{r}
#| label: atdistance
s1@distance
```

Alternately:

```{r}
#| label: distance
distance(s1)
```


## Line passing through the segment

The line that passes through the segment contains information about the segment, such as its slope, intercept, or angle.

To access the line that passes between both points:

```{r}
#| label: segline
s1@line
s1@line@slope
s1@line@intercept
s1@line@angle
```


```{r}
#| label: fig-line
#| fig-cap: Line passing through segment
#| code-fold: true
bp +
  s1@line |> set_props(color = "royalblue4") +
  s1@midpoint(position = c(0, 1))@label(
    polar_just = ob_polar(s1@line@angle + degree(90), 1.5),
    plot_point = TRUE) +
  ob_label(c(equation(s1@line),
             paste0("Distance = ",
                    round(s1@distance, 2))),
           center = midpoint(s1),
           vjust = c(-.2, 1.1),
           angle = s1@line@angle)
```


## Midpoints

By default, the `midpoint` function's `position` argument is .5, which finds the point halfway between the point of a segment:

```{r}
#| label: midpoint
s1@midpoint()
```

```{r}
#| label: fig-midpoint
#| fig-cap: Midpoint of a segment
#| code-fold: true
bp +
  s1 +
  s1@midpoint()@label("Midpoint", hjust = 0, vjust = 1) +
  s1@midpoint(c(0, .5, 1))@label(
    plot_point = TRUE,
    hjust = 1,
    vjust = 0)
```


To find the midpoint 25% of the distance between the endpoints of segment:

```{r}
#| label: midpoint25
s1@midpoint(position = .25)
```


```{r}
#| label: fig-midpoint25
#| fig-cap: Midpoint
#| code-fold: true

bp +
  s1 +
  {p25 <- s1@midpoint(.25)} +
  p25@label(
    label = c(
      p25@auto_label,
      "Midpoint a quarter the<br>way between endpoints"
    ),
    vjust = c(0, 1),
    hjust = c(1, 0)
  ) +
  s1@midpoint(c(0, 1))@label(plot_point = TRUE,
                             hjust = 1,
                             vjust = 0)
```

Multiple midpoints can be specified:

```{r}
#| label: fig-multiplemidpoints
#| fig-cap: Selecting multiple midpoints
bp +
  s1 +
    s1@midpoint(seq(0, 1, .2))
```

A quick way to get the endpoints of a segment is to specify "midpoints" at positions 0 and 1:

```{r}
#| label: fig-endpoints
#| fig-cap: Selecting a segment's endpoints via the `midpoint` property.
bp +
  s1 +
  s1@midpoint(c(0, 1))
```

# Segment chains

If a point object with multiple points is placed in the `p1` slot but the `p2` slot is left empty, a series of segments chained together will be created.

```{r}
#| label: fig-chain
#| fig-cap: Chained segments
bp +
  ob_segment(
    ob_point(x = c(-4, -2, -1, 2, 4),
             y = c(-2, 1, -2, 3, -3)))
```

# Arrowheads and fins

In the `ob_segment` function, the `arrow_head` and `arrow_fins` arguments can be set with any arrowhead function from the [ggarrow](https://teunbrand.github.io/ggarrow/reference/arrow_ornaments.html) or the [arrowheadr](https://wjschne.github.io/arrowheadr/) packages. Any 2-column matrix will also be acceptable, as shown [here](https://teunbrand.github.io/ggarrow/articles/customisation.html). The default arrowhead can be used via the `arrowhead()` function. The size of the arrowhead and fins can also be controlled independently with the `length_head` and `length_fins` arguments.


```{r}
#| label: fig-arrow
#| fig-cap: Segments with arrowheads and fins
bp + ob_segment(p1, p2, 
                arrow_head = arrowhead(), 
                arrow_fins = ggarrow::arrow_fins_feather(), 
                length_head = 5,
                length_fins = 7, 
                linewidth = 2)
```

# Labeling segments

By default, any label is set at the midpoint of the segment (see @fig-segmentlabel). If it is plain text, it will be just above the segment. Its angle will match that of the segment, but this can be turned off by setting `label_sloped` to `FALSE`. 

```{r}
#| label: fig-segmentlabel
#| fig-cap: Segments with labels
bp + ob_segment(p1, p2, label = "My label")
```

As shown @fig-segmentlabel2, the `ob_label` function gives much more control over how and where the label will appear.

```{r}
#| label: fig-segmentlabel2
#| fig-cap: Segments with arrowheads and fins
bp + ob_segment(p1, p2, 
                label = ob_label(
                  label = "My label", 
                  position = .25, 
                  color = "blue", 
                  fill = class_color("blue")@lighten(), 
                  family = "serif", 
                  label.padding = ggplot2::margin(t = 5,r = 4,b = 3,l = 4)))
```


